How access to care shapes disease burden: the current impact of post-exposure prophylaxis and potential for expanded access to prevent human rabies deaths in Madagascar

### 2020-05-12 docs/manuscript.Rmd
fixing how clinics are added
- + various figure fixes
- + updates to man & bib
- + functions to ttimes & fixed out session so that it gives you the max memory used in Mbs

---
title: "How access to care shapes disease burden: the current impact of post-exposure prophylaxis and potential for expanded access to prevent human rabies deaths in Madagascar"
author:
  - Malavika Rajeev:
      email: mrajeev@princeton.edu
      institute: [Princeton]
      correspondence: true
  - Helene Guis:
      institute: [IPMepi, CIRAD]
      correspondence: false
  - Glenn Edosoa:
      institute: [OMS]
      correspondence: false
  - Chantal Hanitriniaina:
      institute: [UoA]
      correspondence: false
  - Soa Fy Andriamandimby:
      institute: [IPMviro]
      correspondence: false
  - Anjasoa Randrianarijaona:
      institute: [IPMepi]
      correspondence: false
  - Reziky Tiandraza Mangahasimbola:
      institute: [IPMepi]
      correspondence: false
  - Ravo Ramiandrasoa:
      institute: [IPMvacc]
      correspondence: false
  - Laurence Baril:
      institute: [IC]
      correspondence: false
  - C.J.E. Metcalf:
      institute: [Princeton]
      correspondence: false
  - Katie Hampson:
      institute: [Glasgow]
      correspondence: false
institute:
  - Princeton: Department of Ecology and Evolutionary Biology, Princeton University, Princeton, United States
  - IPMviro: Virology Unit, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - IPMvacc: Vaccination Center, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - IPMepi: Epidemiology and Clinical Research Unit, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - IC: Institut Pasteur de Cambodge
  - CIRAD: CIRAD, UMR ASTRE, Antananarivo, Madagascar
  - OMS: OMS
  - UoA: Mention Zoologie et Biodiversité Animale, Faculté des Sciences, Université d’Antananarivo, Antananarivo, Madagascar
  - Glasgow: Boyd Orr Centre for Population and Ecosystem Health, Institute of Biodiversity, Animal Health and Comparative Medicine, University of Glasgow, Glasgow, UK
date: ""
link-citations: no
output:
 bookdown::word_document2:
    pandoc_args:
      - --filter=pandoc-crossref
      - --lua-filter=format/scholarly-metadata.lua
      - --lua-filter=format/author-info-blocks.lua
      - --lua-filter=format/pagebreak.lua
      - --lua-filter=format/multiple-bibliographies.lua
      - --reference-doc=format/Reference_Document.docx
 bookdown::html_document2:
    toc: true
    toc_float: true
    pandoc_args:
      - --csl=format/plos-neglected-tropical-diseases.csl
      - --filter=pandoc-crossref
      - --lua-filter=format/scholarly-metadata.lua
      - --lua-filter=format/author-info-blocks.lua
      - --lua-filter=format/pagebreak.lua
      - --lua-filter=format/multiple-bibliographies.lua
keywords: "canine rabies, post-exposure prophylaxis, Gavi, Zero by 30, access to care"
bibliography: ["bib/refs_all.bib", "bib/refs_pkgs.bib"]
# bibliography_main: ["bib/refs_all.bib"]
# bibliography_supp: ["bib/refs_pkgs.bib"]
---

@@

opts_chunk$set(echo = FALSE, warning = FALSE)

@@

# ABSTRACT
Access to care shapes burden of diseases globally. For dog-mediated rabies, post-exposure prophylaxis (PEP) is highly effective at preventing human deaths, however access to PEP is limited in rabies endemic countries. Recently, a target to eliminate deaths due to canine rabies has been set forth by global health partners. In 2018, Gavi added human rabies vaccine to its investment portfolio, a decision which has the potential to greatly reduce rabies deaths. Here, we estimate rabies burden in Madagascar incorporating estimates of geographic access to PEP. We find that travel times to the closest Anti-Rabies Medical Center (ARMC) is a strong predictor of reported bite incidence across the country. We use resulting estimates of bite incidence in an adapted decision tree framework to generate estimates of burden and reporting across the country, and find that spatial access to PEP shapes local burden. We evaluate the impact of expanding access to PEP, using a strategic framework which minimizes travel time for people at the greatest risk of death given a rabies exposure (persons > 3 hrs away from a clinic). Under these assumptions, we estimate that provisioning of PEP to an additional 200 clinics in Madagascar could reduce the burden of rabies deaths by 50%. These results highlight the importance of accounting for access to care in predicting the impacts of interventions sub-nationally and the potential for expanding PEP access strategically to reduce rabies deaths in endemic settings.

@@

A primary challenge is the lack of data on rabies exposures in humans and incidence in animals in most rabies-endemic countries [@scottAddressingDisconnectEstimated2017]. The majority of rabies burden studies at the national level have integrated bite patient data on reported bites into a probability decision tree framework, with the key assumption that overall reported bite incidence (i.e. both bites due to non-rabid and rabid animals) are proportional to rabies incidence (i.e. the more bites reported in a location, the higher the incidence of *rabies* exposures), and that reporting is uniform across space [@hampsonEstimatingGlobalBurden2015; @knobelReevaluatingBurdenRabies2005; @rajeevHealthcareUtilizationProvisioning2018]. If applied subnationally, these assumptions would likely underestimate rabies deaths in places with poor access to PEP and may overestimate rabies deaths in places with better access to PEP.

@@

## Estimating travel time to closest ARMC {#methods-spatial}
To estimate travel times to the nearest clinic at the scale of administrative units, we used two raster inputs: 1) the friction surface from the Malaria Atlas Project [@weissGlobalMapTravel2018] at an ~1 km^2 scale ([Fig S1.1A](#figS1.1)) and 2) the population estimates which we resampled to the friction surface from the the 2015 UN adjusted population projections from World Pop (@linardPopulationDistributionSettlement2012a, originally at an ~100m^2 resolution, [Fig S1.1B](#figS1.1)).  
Using the approach described in @weissGlobalMapTravel2018, we used GPS points of ARMC to estimate the travel time to the nearest clinic provisioning PEP for the country of Madagascar at an approximately 1 x 1 km scale. We then extracted the mean of travel times for each administrative unit (district commune), as well as the mean of travel times weighted by the population in that grid cell. We also compared these metrics to euclidean distance, but taking the minimum distance from the administrative unit centroid to an ARMC. We used administrative shapefiles from the UN Office for the Coordination of Humanitarian Affairs.

@@

## Estimating average annual bite incidence {#methods-bitedata}
We used two datasets on bite patients reporting to ARMC:
1) A national database of bite patient forms submitted to IPM from ARMC across the country between 2014 - 2017. These were individual patient data forms that were submitted from clinics as frequently as monthly to annually that included data on the date of reporting and were resolved to the district level (i.e. the district where the patient resided)
2) `r max(mora_bites$nobs, na.rm = TRUE)` months of data (between October 2016 and June 2019) on patients reporting to the Moramanga ARMC resolved to the commune level (the administrative level below the district). We calculated average annual bite incidence to the commune level by taking the monthly average of reported bites and multiplying by 12.

For the national data, some clinics did not report any data, and for others there were substantial periods, months to even a whole year, when no patient forms were submitted. To correct for under-submission of forms, we attempted to identify what proportion of data were missing each year by excluding any periods of time for which there were 15 consecutive days where zero patient records were submitted ([Figure S2.1A](figS2.1). We validated this cut-off of consecutive days by comparing estimated vial demand given the total reported bites (corrected for under-submission) for the years 2014-2016 to actual vials provisioned to clinics over this same period. We did not have the date or month when vials were provisioned, so we compared the totals across this three year period.

Vial demand was simulated under simplified assumptions of PEP administration and adherence[@rajeevHealthcareUtilizationProvisioning2018], assuming that patients reported randomly across the year and then completed doses 3, 7, and 28 days subsequent from the date reported. For this analysis, we look at the range of vial estimates assuming that all patients complete 3 vs. 4 doses total. During this period, the Thai Red Cross Intradermal regimen was used across Madagascar, with 0.2 mL administered per patient at each visit. Vials can be shared within a day between two patients, resulting in 0.1 background wastage per vial shared + any additional wasteage from unused doses discarded at the end of the day. While these simplifying assumptions do not capture variation in adherence, completion, delays, or errors in administration, PEP completion is overall high in Madagascar and in other settings where PEP is provided at no direct costs to patients, so our estimates at 3 doses per patient vs. 4 are likely encompassing much of this variation at the annual level.

We then divided the total number of bites reported to each clinic from each clinic in a given year by the undersubmission rate for eachclinic. Due to the variation in submission rates across time, we chose to take the average of the annual bite incidence estimates aggregated to the district level. The data from the Moramanga district were completely reported across the `r max(mora_bites$nobs, na.rm = TRUE)` month period.

@@

## Modeling reported bites as a function of travel time to closest ARMC {#methods-bitemods}
We modeled the number of reported bites as a function of travel time ($T$) in a poisson regression framework as follows: 
$$ \mu_i = e^{(\beta_{t}T_i \ + \ \beta_0 \ )}P_i$$
$$ y_i = Poisson(\mu_i) $$
where $y_i$ is our data on average number of bites reported to a clinic annually and $\mu_i$ is the expected number of bite patients presenting at the ARMC as a function of travel time to the ARMC ($T_i$) and the human population size ($P_i$) (an offset which scales the predicted number of bites) for a given source location (district or commune). We fit this model to both the national data resolved to the district level and the Moramanga data resolved to the commune level in a Bayesian regression framework via MCMC using the R package ‘rjags’. In order to more directly compare estimates between these two datasets, we also modeled the national data with a commune-level travel time covariate ($T_{j}$) so that:

$$\mu_i = \sum \limits_{j=1}^j e^{(\beta_{t}T_j\ + \ \beta_0 _j)}P_j$$

@@

For the models fitted to the national data, we also attempted to capture variation between clinics by modeling a random catchment effect of the form: $B_{0,k} \sim norm(\mu,\ \sigma)$) where $\mu$ is the mean and $\sigma$ is standard deviation and $B_{0,k}$ is the catchment level intercept). We tested whether the catchment random effect was capturing overdispersion rather than any catchment specific effects by comparing the ratio of the sum of squared residuals to the expected $\chi^2_{n - k}$ distribution given a poisson error structure [@gelman2006data].

We compared alternative model structures for predicting reported bites, specifically by including human population size as a predictor of reported bites (addPop = population size as additional covariate, onlyPop = population as only covariate, flatPop = population as offset in model), as well as by intercept type ( random = random intercept by catchment, fixed = a single fixed intercept was estimated) for the national data (see main Methods for specific formulations). We used priors centered around zero (*norm(0, 10-6)*) except for the precision parameter for which we used an uniform distribution (*unif(0, 100)*).\

The Moramanga data were not fit to models with a random catchment effect, as these were data from a single catchment. We calculated Deviance Information Criterion (DIC, a metric of model fit to data) for each candidate model as well as potential scale reduction factors (psrf, a metric of model convergence) for each covariate in the model and the multivariate psrf for the whole model (cite Gelman book).

To test our models predictions further, we used parameter estimates from each model to generate predictions for data that were not used to fit the model (i.e. we used the estimates from models fitted to the national data to predict the Moramanga data, and used estimates from models fitted to the Moramanga data to predict the national data). Finally, to check how correcting for incomplete submission of forms affected our modeling results, we fitted these same models to the raw data uncorrected for submission (i.e. assuming that data was completely reported) and with a lower cut-off for identifying periods of undersubmission (7 days vs. 15 days, resulting in higher estimates of undersubmission and accordingly bite incidence).

## Modeling the burden of human rabies deaths sub-nationally {#methods-burden}

We estimate deaths due to rabies as a function of the number of bites predicted by our model and estimates of endemic rabies exposure incidence using an adapted decision tree framework (Fig 2). As with previous decision tree studies [@cleavelandEstimatingHumanRabies2002; @hampsonEstimatingGlobalBurden2015], to model uncertainty in parameter estimates we used triangular distributions for two key parameters: $E_i$, the annual human exposure incidence in a given administrative unit, and $p_rabid$, the proportion of reported bites that are rabies exposures. The range for these distributions was taken from a previous study in Moramanga and also informed by recent methods for estimating underlying rabies incidence in an endemic setting when data is lacking([@whorabiesmodelingconsortiumPotentialImpactImproved2019]).

Specifically: annual rabies incidence of 1% in the dog population in an endemic setting with no mass dog vaccination . To generate $E_i$, the human exposure incidence in a given administrative unit (where $i$ is the index of the commune or district) based on the model, we multiply this baseline incidence by the estimated dog population and the exposure rate per rabid dog ($p_{exp}$). We estimate the dog population given a human:dog ratio (HDR) of 7 to generate our maximum expected incidence and an HDR of 35 for our minimum expected incidence, giving a range of 11.1 ($E_{min}$) - 55.7 ($E_{max}$) exposures per 100,000 persons. As there is little data on dog populations in Madagascar [@andriamandimbySurveillanceControlRabies2013], this range encompasses both a wide range of observed HDRs [@knobelReevaluatingBurdenRabies2005; @sambo2018estimating] and generates similar human exposure incidences as observed previously across Africa [cite here!] . $E_i$ is then drawn from a uniform distribution between the minimum and maximum expected number of human exposures for each location.

@@

| **Parameter**   | **Value**        | **Description**	                 | **Source**|
|:----------------|:----------------:|:----------------------------------|:---------:|
| $E_i$ |	Triangular(a = 15.6 b = 76, c = 42)| Annual exposures per 100,000 persons	| [@whorabiesmodelingconsortiumPotentialImpactImproved2019; @rajeevHealthcareUtilizationProvisioning2018], see S4|
| $p_{rabid}$	| Triangular(a = 0.2 b = 0.6, c = 0.4) |	Proportion of reported bites that are rabies exposures | [@rajeevHealthcareUtilizationProvisioning2018] |
| $\rho_{max}$ | 0.98	| The maximum reporting possible for any location	Data from Moramanga Ville, Moramanga District | [@rajeevHealthcareUtilizationProvisioning2018] |
| $p_{death}$ |	0.16 | The probability of death given a rabies exposure	| [@changaluchaNeedImproveAccess2018]|

@@

```{r ncsbs}
## candidate points
csbs <- read.csv(here("data/raw/csbs.csv"), stringsAsFactors = FALSE)
csbs %>% 
  filter(type == "CSB2", genre_fs != "Priv", type_fs != "Health Post") %>%
  dplyr::select(CTAR = nom_fs, X_COORD = ycoor, Y_COORD = xcoor) -> csbs
```

@@

Starting with the current locations, we added one clinic at a time, calculating the proportion of people living < 3 hours away from any clinic for the country. We added the clinic which minimized this metric and then repeated the process iteratively, ranking clinics and adding the top clinic sequentially. In this iterative process, we also exclude any clinics that once added only reduced travel times below the 3 hour threshold for less than 0.01% of the population (< `r round(sum(mada_districts$pop)*0.0001, -2)` persons). We also explore the scenario in which all CSB II in the country provision PEP to get an estimate of the maximum impact of expanding PEP provisioning. Resulting travel time estimates and locations of when and where clinics were added are shown in [Supplementary section S5](#S5). Finally, we estimate the incremental reduction in burden as each clinic is added using the burden framework described in the previous section.

We estimate how expansion to additional clinics may impact vaccine demand. Specifically, we use our model-predicted estimates of annual bites at the district and commune level and then aggregate these to the clinic level. As clinics are added, multiple clinics may serve a single district or commune (Fig TO CHECK S5.3B). Therefore, we allocated bites to each clinic according to the proportion of the population in each administrative unit for which it was the closest (Fig CHECK maps). For each clinic, we simulate throughput by randomly assigning patient presentation dates, and then assume perfect compliance (i.e. all patients reporting 3 and 7 days after their first dose) to generate dates of subsequent doses. We use these dates to estimate vial usage given routine vial sharing practices in Madagascar (i.e. each vial is split between two patients, and vials are discarded at the end of each day) and the new abridged intradermal regimen recommended by WHO (2 injection sites on day 0, 3, and 7, [@tarantolaIntradermalRabiesPostexposure2018]). We assume that vial sharing is functioning perfectly, with a background wastage of 0.1 mL per vial, but not accounting for wastage due to other reasons (i.e. wastage due to injection error). We take the mean of 1000 simulations of throughput for each catchment and set of clinics.

## Sensitivity of burden estimates to decision tree parameters {#methods-sensitivity}
To test the effect of our model assumptions on estimates of rabies burden and vial demand, we did a univariate sensitivity analyses of both parameters from the models of bite incidence and models of burden. Additionally, we also examine the systematic variation in human rabies exposure incidence with human population (a potential proxy for changes in the dog population, see section S6) by looking at how estimates of burden change if rabies incidence were to scale positively or negatively with human population size. 
We also generate scaling factors to look at how systematic variation in rabies incidence with human population size could affect our estimates of burden (Fig S6.1). Specifically, if we believe that human to dog ratios are positively correlated with human populations (i.e. if dog ownership/populations are higher in more populated, urban areas), then we might expect higher rabies exposure incidence within our range as population size increases. Alternatively, if we believe that the human to dog ratio is inversely correlated with human population size (i.e. if dog ownership is more common in less populated, rural areas), we would expect rabies exposure incidence within our range to scale negatively with human population size.

## Data and code 
All analyses were done in `r sessionInfo()$R.version$version.string` [@RLang2019] and using the packages listed in the [supplementary references]{#srefs}. All processed data, code, and outputs are archived at this zenodo repository and maintained at this github repo. The raw bite patient data at the national level is maintained by IPM. This database was last queried on September 19, 2019 for these analyses. Anonymized raw data from this database are available from IPM following insitutional data transfer protocols. Anonymized raw data collected from the Moramanga District were retrieved from the Wise Monkey Portal (http://www.wisemonkeyfoundation.org/) on the same date and are shared in the archived repository.

>>>>

---
title: "How access to care shapes disease burden: the current impact of post-exposure prophylaxis and potential for expanded access to prevent human rabies deaths in Madagascar"
author:
  - Malavika Rajeev:
      email: mrajeev@princeton.edu
      institute: [Princeton]
      correspondence: true
  - Helene Guis:
      institute: [IPMepi, CIRAD]
      correspondence: false
  - Glenn Edosoa:
      institute: [OMS]
      correspondence: false
  - Chantal Hanitriniaina:
      institute: [UoA]
      correspondence: false
  - Anjasoa Randrianarijaona:
      institute: [IPMepi]
      correspondence: false
  - Reziky Tiandraza Mangahasimbola:
      institute: [IPMepi]
      correspondence: false
  - Fleur Hierink:
      institute: [Geneva]
      correspondence: false
  - Ravo Ramiandrasoa:
      institute: [IPMvacc]
      correspondence: false
  - Jean-Michel Heraud:
      institute: [IPMviro]
      correspondence: false
  - Soa Fy Andriamandimby:
      institute: [IPMviro]
      correspondence: false
  - Laurence Baril:
      institute: [IPCambodia]
      correspondence: false
  - C.J.E. Metcalf:
      institute: [Princeton]
      correspondence: false
  - Katie Hampson:
      institute: [Glasgow]
      correspondence: false
institute:
  - Princeton: Department of Ecology and Evolutionary Biology, Princeton University, Princeton, United States
  - IPMviro: Virology Unit, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - IPMvacc: Vaccination Center, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - IPMepi: Epidemiology and Clinical Research Unit, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - IC: Institut Pasteur de Cambodge
  - CIRAD: CIRAD, UMR ASTRE, Antananarivo, Madagascar
  - OMS: OMS
  - UoA: Mention Zoologie et Biodiversité Animale, Faculté des Sciences, Université d’Antananarivo, Antananarivo, Madagascar
  - Glasgow: Boyd Orr Centre for Population and Ecosystem Health, Institute of Biodiversity, Animal Health and Comparative Medicine, University of Glasgow, Glasgow, UK
  - Geneva: University of Geneva
date: ""
link-citations: no
output:
 bookdown::word_document2:
    pandoc_args:
      - --filter=pandoc-crossref
      - --lua-filter=format/scholarly-metadata.lua
      - --lua-filter=format/author-info-blocks.lua
      - --lua-filter=format/pagebreak.lua
      - --lua-filter=format/multiple-bibliographies.lua
      - --reference-doc=format/Reference_Document.docx
 bookdown::html_document2:
    toc: true
    toc_float: true
    pandoc_args:
      - --csl=format/plos-neglected-tropical-diseases.csl
      - --filter=pandoc-crossref
      - --lua-filter=format/scholarly-metadata.lua
      - --lua-filter=format/author-info-blocks.lua
      - --lua-filter=format/pagebreak.lua
      - --lua-filter=format/multiple-bibliographies.lua
keywords: "canine rabies, post-exposure prophylaxis, Gavi, Zero by 30, access to care"
bibliography: ["bib/refs_all.bib", "bib/refs_pkgs.bib"]
# bibliography_main: ["bib/refs_all.bib"]
# bibliography_supp: ["bib/refs_pkgs.bib"]
---

@@

opts_chunk$set(echo = FALSE, warning = FALSE, include = FALSE)

@@

# ABSTRACT
Access to care shapes burden of diseases globally. For dog-mediated rabies, post-exposure prophylaxis (PEP) is highly effective at preventing human deaths, however access to PEP is limited in rabies endemic countries. Recently, a target to eliminate deaths due to canine rabies has been set forth by global health partners. In 2018, Gavi added human rabies vaccine to its investment portfolio, a decision which has the potential to greatly reduce rabies deaths. Here, we estimate rabies burden in Madagascar incorporating estimates of geographic access to PEP.

Go in order.

We find that travel times to the closest Anti-Rabies Medical Center (ARMC) is a strong predictor of reported bite incidence across the country. We use resulting estimates of bite incidence in an adapted decision tree framework to generate estimates of burden and reporting across the country, and find that spatial access to PEP shapes local burden.

We evaluate the impact of expanding access to PEP, using a strategic framework which minimizes travel time for people at the greatest risk of death given a rabies exposure (persons > 3 hrs away from a clinic). Under these assumptions, we estimate that provisioning of PEP to an additional 200 clinics in Madagascar could reduce the burden of rabies deaths by 50%.

However our estimates are dependent on blah blah blah.

These results highlight the importance of accounting for access to care in predicting the impacts of interventions sub-nationally and the potential for expanding PEP access strategically to reduce rabies deaths in endemic settings.

@@

A primary challenge is the lack of data on rabies exposures in humans and incidence in animals in most rabies-endemic countries [@scottAddressingDisconnectEstimated2017]. The majority of rabies burden studies at the national level have integrated bite patient data on reported bites into a probability decision tree framework, with the key assumption that overall reported bite incidence (i.e. both bites due to non-rabid and rabid animals) are proportional to rabies incidence (i.e. the more bites reported in a location, the higher the incidence of rabies exposures), and that reporting is uniform across space [@hampsonEstimatingGlobalBurden2015; @knobelReevaluatingBurdenRabies2005; @rajeevHealthcareUtilizationProvisioning2018]. If applied subnationally, these assumptions would likely underestimate rabies deaths in places with poor access to PEP and may overestimate rabies deaths in places with better access to PEP.

@@

## Estimating travel time to closest ARMC {#methods-spatial}
To estimate travel times to the nearest clinic at the scale of administrative units, we used two raster inputs: 1) the friction surface from the Malaria Atlas Project [@weissGlobalMapTravel2018] at an ~1 km^2 scale ([Fig S1.1A](#figS1.1)) and 2) the population estimates which we resampled to the friction surface from the the 2015 UN adjusted population projections from World Pop (@linardPopulationDistributionSettlement2012a, originally at an ~100m^2 resolution, [Fig S1.1B](#figS1.1)).

Using the approach described in @weissGlobalMapTravel2018, we used GPS points of ARMC to estimate the travel time to the nearest ARMC (i.e. a clinic provisioning PEP) for the country of Madagascar at an approximately 1 x 1 km scale. We then extracted the mean of travel times at the district and commune (the administrative unit below the district) level, as well as the mean of travel times weighted by the population in that grid cell. We also compared these metrics to euclidean distance, but taking the minimum distance from the administrative unit centroid to an ARMC. We used administrative shapefiles from the UN Office for the Coordination of Humanitarian Affairs.

@@

## Estimating average annual bite incidence {#methods-bitedata}
We used two datasets on bite patients reporting to ARMC:

1) A national database of bite patient forms submitted to IPM from ARMC across the country between 2014 - 2017. These were individual patient data forms that were submitted from clinics as frequently as monthly to annually, which included data on the date of reporting and were resolved to the district level (i.e. the district where the patient resided).
2) `r max(mora_bites$nobs, na.rm = TRUE)` months of data (between October 2016 and June 2019) on patients reporting to the Moramanga ARMC resolved to the commune level. We calculated average annual bite incidence to the commune level by taking the monthly average of reported bites and multiplying by 12.

The data from the Moramanga district were completely reported across the `r max(mora_bites$nobs, na.rm = TRUE)` month period. However, for the national data, some clinics did not report any data, and for others there were substantial periods, months to even a whole year, when no patient forms were submitted. To correct for under-submission of forms, we attempted to identify what proportion of data were missing each year by excluding any periods of time for which there were 15 consecutive days where zero patient records were submitted ([Figure S2.1A](figS2.1). We then divided the total number of bites reported to each clinic from each clinic in a given year by the estimated proportion of forms which were not submitted (undersubmission) for each clinic. Due to the variation in submission of forms across time, we chose to take the average of the annual bite incidence estimates aggregated to the district level.

We validated this approach to estimating form undersubmission by comparing estimated vial demand given the total reported bites (corrected for under-submission) for the years 2014-2017 to actual vials provisioned to clinics over this same period. Vial demand was simulated under simplified assumptions of PEP administration and adherence[@rajeevHealthcareUtilizationProvisioning2018], assuming that patients reported randomly across the year and then completed doses 3, 7, and 28 days subsequent from the date reported. For this analysis, we look at the range of vial estimates assuming that all patients complete 3 vs. 4 doses total. During this period, the Thai Red Cross Intradermal regimen was used across Madagascar, with 0.2 mL administered per patient at each visit. Vials can be shared within a day between two patients, resulting in 0.1 background wastage per vial shared, plus any additional wasteage from unused doses discarded at the end of the day. We did not have the date or month when vials were provisioned, so we compared the totals across this four year period.

@@

## Modeling reported bites as a function of travel time to closest ARMC {#methods-bitemods}
We modeled the number of reported bites as a function of travel time ($T$) in a poisson regression framework as follows: 
$$ \mu_i = e^{(\beta_{t}T_i \ + \ \beta_0 \ )}P_i$$
$$ y_i = Poisson(\mu_i) $$
where $y_i$ is our data on average number of bites reported to a clinic annually and $\mu_i$ is the expected number of bite patients presenting at the ARMC as a function of travel time to the ARMC ($T_i$) and the human population size ($P_i$) (an offset which scales the incidence to expected number of bites) for a given source location (district or commune). We fit this model to both the national data resolved to the district level and the Moramanga data resolved to the commune level. In order to more directly compare estimates between these two datasets, we also modeled the national data with a commune-level travel time covariate ($T_{j}$) so that:

$$\mu_i = \sum \limits_{j=1}^j e^{(\beta_{t}T_j\ + \ \beta_0_j)}P_j$$

@@

For the models fitted to the national data, we also attempted to capture variation between clinics by modeling a random catchment effect of the form: $B_{0,k} \sim norm(\mu,\ \sigma_0)$) where $\mu$ is the mean and $\sigma_0$ is standard deviation and $B_{0,k}$ is the catchment level intercept). The Moramanga data were not fit to models with a random catchment effect, as these were data from a single catchment.

We tested whether the catchment random effect was capturing overdispersion rather than any catchment specific effects by extending these models to incorporate an overdispersion parameter, $\sigma_e$ [@gelman2006data]:

$$ \mu_i = e^{(\sum \limits_{j=1}^j \beta_jX_j + \sigma_e)}P_i$$
where $\sum \limits_{j=1}^j \beta_jX_j$ is the sum of the all parameters and for a given model.

We fit all models in a Bayesian regression framework via MCMC using the R package ‘rjags’. We used a weakly informative prior for the intercept of the models centered around the mean of the bite incidence for the given data set, and for all other parameters we centered the priors around zero ($norm(0, 10-6)$). For the variance terms ($\sigma_0$ and $\sigma_e$), we set uniform priors ($unif(0, 10)$). We calculated the Deviance Information Criterion (DIC, a metric of model fit to data) for each candidate model as well as potential scale reduction factors (psrf, a metric of model convergence) for each covariate in the model and the multivariate psrf for the whole model [@gelman2006data].

To test how well our models predicted the data, we sampled parameter estimates from the posterior distribution for each model to generate predictions to the data. In addition, we used the models to predict the data that were not used to fit the model (i.e. we used the estimates from models fitted to the national data to predict the Moramanga data, and used estimates from models fitted to the Moramanga data to predict the national data). Finally, to check how correcting for incomplete submission of forms affected our modeling results, we fitted our final models to the raw data uncorrected for submission (i.e. assuming that data was completely reported) and with a lower cut-off for identifying periods of undersubmission (7 days vs. 15 days, resulting in higher estimates of undersubmission and accordingly bite incidence).

## Modeling the burden of human rabies deaths sub-nationally {#methods-burden}
We estimate deaths due to rabies as a function of the number of bites predicted by our model and estimates of endemic rabies exposure incidence using an adapted decision tree framework (Fig 1). To model uncertainty in parameter estimates we used triangular distributions, as with previous decision tree studies [@cleavelandEstimatingHumanRabies2002; @hampsonEstimatingGlobalBurden2015], for two key parameters: $E_i$, the annual human exposure incidence in a given administrative unit, and $p_rabid$, the proportion of reported bites that are rabies exposures. The range for these distributions was taken from a previous study in Moramanga and also informed by recent methods for estimating underlying rabies incidence in an endemic setting when data is lacking([@whorabiesmodelingconsortiumPotentialImpactImproved2019]).

For $E_i$, we center the distribution at the lower end of our estimate of exposure incidence from the Moramanga District (42 exposures/100,000 persons). The range is estimated using the approach used in [@whorabiesmodelingconsortiumPotentialImpactImproved2019] where rabies incidence in the dog population is estimated to be approximately 1% annually in the absence of dog vaccination, and that on average a rabid dog exposes 0.39 persons. These assumptions are derived from modeling and long term contact tracing data from the Serengeti District, Tanzania [@hampsonRabiesExposuresPostExposure2008; @hampsonSurveillanceEstablishElimination2016; @whorabiesmodelingconsortiumPotentialImpactImproved2019]. We then take the range of exposure incidence as the probability of exposure multiplied by the number of expected rabies cases in dogs assuming 1% incidence and the estimate of the dog population using a range of human to dog ratios (HDRs) of 5 (upper estimate) and 25 (lower estimate). As there is little data on dog populations in Madagascar [@andriamandimbySurveillanceControlRabies2013], this range of HDRs encompasses a wide range of observed HDRs across Africa [@knobelReevaluatingBurdenRabies2005; @sambo2018estimating], as well as recent subnational estimates from Madagascar [@leblancclaireRabiesMadagascarThreepronged2019], and also generates similar human exposure incidences as observed previously across Africa [CITE!!]. See figure S4 for further details.

@@

| **Parameter**   | **Value**        | **Description**	                 | **Source**|
|:----------------|:----------------:|:----------------------------------|:---------:|
| $E_i$ |	Triangular(a = 15, b = 76, c = 42)| Annual exposures per 100,000 persons	| [@whorabiesmodelingconsortiumPotentialImpactImproved2019; @rajeevHealthcareUtilizationProvisioning2018], see Fig S4.1|
| $p_{rabid}$	| Triangular(a = 0.2, b = 0.6, c = 0.4) |	Proportion of reported bites that are rabies exposures | [@rajeevHealthcareUtilizationProvisioning2018] |
| $\rho_{max}$ | 0.98	| The maximum reporting possible for any location	Data from Moramanga Ville, Moramanga District | [@rajeevHealthcareUtilizationProvisioning2018] |
| $p_{death}$ |	0.16 | The probability of death given a rabies exposure	| [@changaluchaNeedImproveAccess2018]|

@@

```{r ncsbs}
## candidate points
csbs <- read.csv(here("data/processed/clinics/csb2.csv"), stringsAsFactors = FALSE)
clin_per_comm <- read.csv(here("data/processed/clinics/clinic_per_comm.csv"), 
                          stringsAsFactors = FALSE)
```

@@

Starting with the current locations, we added one clinic at a time, calculating the proportion of people living < 3 hours away from any clinic for the country. We added the clinic which minimized this metric and then repeated the process iteratively, ranking clinics and adding the top clinic sequentially. In this iterative process, we also exclude any clinics that once added only reduced travel times below the 3 hour threshold for less than 0.01% of the population (< `r round(sum(mada_districts$pop)*0.0001, -2)` persons). Finally, we estimate the incremental reduction in burden as each clinic is added using the burden framework described in the previous section.

We also explore three additional scenarios: 
1) Each district has one clinic (the CSB II in the highest population density grid cell compared to other potential clinics) provision PEP (N = 114 clinics). 
2) Each commune has one clinic (the CSB II, or when there are no CSB II in a commune the health post or dispensary, in the highest density grid cell compared to other potential clinics) provision PEP (N = `r nrow(clinic_per_comm)`. Note that not all communes (N = `r length(mada_communes)`) have a clinic).
3) All CSB II in the country provision PEP (N = `r nrow(csbs)` clinics).

We estimate how expansion to additional clinics may impact vaccine demand. Specifically, we use our model-predicted estimates of annual bites at the district and commune level and then aggregate these to the clinic level. As clinics are added, multiple clinics may serve a single district or commune (S5.4). Therefore, we allocated bites to each clinic according to the proportion of the population in each administrative unit for which it was the closest. For each clinic, we simulate throughput by randomly assigning patient presentation dates, and then assume perfect compliance (i.e. all patients reporting 3 and 7 days after their first dose) to generate dates of subsequent doses. We use these dates to estimate vial usage given routine vial sharing practices in Madagascar as described previously, but under the assumption of adoption of the new abridged intradermal regimen recommended by WHO (2 injection sites on day 0, 3, and 7, [@tarantolaIntradermalRabiesPostexposure2018]). We take the mean of 1000 simulations of throughput for each catchment and set of clinics.

## Sensitivity analysis {#methods-sensitivity}
To test the effect of our model assumptions on estimates of rabies burden and vial demand, we did a univariate sensitivity analyses of both parameters from the models of bite incidence and the decision tree parameters. For each parameter in the decision tree, we simulated predcitions fixing estimates to the upper and lower range of each parameter, while keeping the other parameters unchanged (Table S6.1). We did the same for parameter estimates from the models of reported bite incidence, but rather fixed these estimates at the upper and lower 95% quantiles of their posterior distributions (Table S6.2). We also look at how systematic variation in rabies incidence with human population size could affect our estimates of burden. Specifically, if we believe that human to dog ratios are positively correlated with human populations (i.e. if dog ownership/populations are higher in more populated, urban areas), then we might expect higher rabies exposure incidence within our range as population size increases. Alternatively, if we believe that the human to dog ratio is inversely correlated with human population size (i.e. if dog ownership is more common in less populated, rural areas), we would expect rabies exposure incidence within our range to scale negatively with human population size. We use scaling factors which scale incidence either positively or negatively with the population sizes observed at the district and commune levels, but are still constrained within the range of exposure incidence used in the main analyses (15.6 - 76 exposures per 100,000 persons, see Fig S6.4).

## Data and code 
All analyses were done in `r sessionInfo()$R.version$version.string` [@RLang2019] and using the packages listed in the [supplementary references]{#srefs}. All processed data, code, and outputs are archived at this zenodo repository (to be archived upon submission) and maintained at this github repository. The raw bite patient data at the national level are maintained in a database by IPM. This database was last queried on September 19, 2019 for these analyses. Anonymized raw data from this database are available from IPM following insitutional data transfer protocols. Anonymized raw data collected from the Moramanga District were retrieved from the Wise Monkey Portal (http://www.wisemonkeyfoundation.org/) on the same date and are shared in the archived repository.
